---
title: "Assignment3"
author: "RF"
date: "3/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Multivariate linear models

In this assignment you will investigate the relation between different aspects of IQ and symptom severity in ASD (as measured by ADOS). The assignment is meant to make you practice on linear models, multiple predictors and interactions.

The data you will use is the outcome of the psychological testing of the children you analyzed in methods 3.
Data: https://www.dropbox.com/s/hom2qnmp9hyfwmu/Assignment3Data.csv?dl=0
The data consists of ChildID, gender, age, diagnosis (ASD), symptom severity (ADOS), several aspects of IQ (NonVerbal/Perceptual,Verbal,Social) and the visit at which the tests were administered. The questions should be answered by focusing on visit 1.

```{r}
library(rethinking)
library(pacman)
library(magrittr)
library(ggplot2)
```

```{r}
d = read.csv("Assignment3Data.csv")

# We need only visit 1 and ASDs 
d_sub = subset(d, Visit=="1" & ASD =="1")

# There is also one child with ADOS=0 - what happends if we take them out?
#d_sub2 = subset(d, Visit=="1" & ASD =="1" & ADOS >"0")

# TDs and ASDs, visit 1
d2 = subset(d, Visit=="1")
d2 <- d2[ complete.cases(d2) , ]

# Only TDs
d3 = subset(d, Visit=="1" & ASD =="0")
d3 <- d3[ complete.cases(d3) , ]
```


The questions you will have to answer are:

1. Assess the relation between symptom severity and IQ (focus on visit 1 and children with ASD) and report the model, a plot of the model, a couple of lines describing the quality of the model and interpreting the results. P.S. Should you scale? P.P.S. Don't forget to motivate your priors. P.P.P.S. At least one plot for results and a plot for quality of each model (here and in the next questions) would be appreciated.
1.1. Non Verbal IQ and ADOS
1.2. Verbal IQ and ADOS
1.3. Social IQ and ADOS

```{r}
# Rescale 
  # List of variables
rescalelist = c("NonVerbalIQ", "VerbalIQ", "SocialIQ", "ADOS") 
 
# ASDs only 
d_sub.s = d_sub[, colnames(d_sub) %in% rescalelist] %>% #select rows to rescale 
  lapply(. , function(x) scale(x, center = mean(x, na.rm =T), scale = sd(x, na.rm = T))) %>%  #adding mean and sd manually to add the na.rm = T to removes NA's
  cbind(. , d_sub[,! colnames(d_sub) %in% rescalelist]) #bind with remaining rows

# Everyone
d2.s = d2[, colnames(d2) %in% rescalelist] %>% #select rows to rescale 
  lapply(. , function(x) scale(x, center = mean(x, na.rm =T), scale = sd(x, na.rm = T))) %>%  #adding mean and sd manually to add the na.rm = T to removes NA's
  cbind(. , d2[,! colnames(d2) %in% rescalelist]) #bind with remaining rows

# TD only
d3.s = d3[, colnames(d3) %in% rescalelist] %>% #select rows to rescale 
  lapply(. , function(x) scale(x, center = mean(x, na.rm =T), scale = sd(x, na.rm = T))) %>%  #adding mean and sd manually to add the na.rm = T to removes NA's
  cbind(. , d3[,! colnames(d3) %in% rescalelist]) #bind with remaining rows
```


```{r}
# First, plot IQ as a function of ADOS
plot(d_sub$NonVerbalIQ ~ d_sub$ADOS)
plot(d_sub$VerbalIQ ~ d_sub$ADOS)
plot(d_sub$SocialIQ ~ d_sub$ADOS)

# Plot the distribution of IQ
dens(d_sub$NonVerbalIQ)
dens(d_sub$VerbalIQ)
dens(d_sub$SocialIQ)

#----Not using this rn----
# Plot
# Superimpose the MAP values for mean height over the actual data
plot(ADOS ~ NonVerbalIQ , data=d_sub.s )
abline( a=coef(m1)["a"] , b=coef(m1)["b"] )
# Extract samples from the model
post <- extract.samples(m1)
# Extract first 10 cases and re-estimate the model
N <- 10
dN <- d_sub.s[ 1:N , ]
mN <- map(
  alist(
    ADOS ~ dnorm( mu , sigma ) ,
    mu <- a + b*NonVerbalIQ ,
    a ~ dnorm(0,0.5) ,
    b ~ dnorm( 0 , 0.5 ) ,
    sigma ~ dunif( 0 , 2 )
  ) , data=dN )
# Extract 20 samples from the posterior
post <- extract.samples( mN , n=20 )
# Display raw data and sample size
plot( dN$NonVerbalIQ, dN$ADOS,
      xlim=range(d_sub.s$NonVerbalIQ) , ylim=range(d_sub.s$ADOS) ,
      col=rangi2 , xlab="NonVerbalIQ" , ylab="ADOS" )
mtext(concat("N = ",N))
# Plot the lines, with transparency
for ( i in 1:20 )
  abline( a=post$a[i] , b=post$b[i] , col=col.alpha("black",0.3) )
#------------

## NonVerbalIQ
m1 <- map(
  alist(
    ADOS ~ dnorm(mu, sigma),
    mu <- a + b*NonVerbalIQ,
    a ~ dnorm(0, 0.5),
    b ~ dnorm(0,0.5),    
    sigma ~ dunif(0,2)   
  ),
  data = d_sub.s)

precis(m1)

# Plots
precis_plot(precis(m1))

# compute percentile interval of mean
nvIQ.seq <- seq( from=-3 , to=3 , length.out=30 )
mu <- link( m1 , data=data.frame(NonVerbalIQ = nvIQ.seq) )
mu.PI <- apply( mu , 2 , PI )
# plot it all
plot( ADOS ~ NonVerbalIQ , data=d_sub.s , col=rangi2 )
abline( m1 )
shade( mu.PI , nvIQ.seq )


## VerbalIQ
m2 <- map(
  alist(
    ADOS ~ dnorm(mu, sigma),
    mu <- a + b*VerbalIQ,
    a ~ dnorm(0, 0.5),
    b ~ dnorm(0,0.5),    
    sigma ~ dunif(0,2)   
  ),
  data = d_sub.s)

precis(m2)

# Plots
precis_plot(precis(m2))

# compute percentile interval of mean
vIQ.seq <- seq( from=-3 , to=3 , length.out=30 )
mu <- link( m2 , data=data.frame(VerbalIQ = vIQ.seq) )
mu.PI <- apply( mu , 2 , PI )
# plot it all
plot( ADOS ~ VerbalIQ , data=d_sub.s , col=rangi2 )
abline( m2 )
shade( mu.PI , vIQ.seq )


## SocialIQ
m3 <- map(
  alist(
    ADOS ~ dnorm(mu, sigma),
    mu <- a + b*SocialIQ,
    a ~ dnorm(0, 0.5),
    b ~ dnorm(0,0.5),   
    sigma ~ dunif(0,2)   
  ),
  data = d_sub.s)

precis(m3)

# Plots
precis_plot(precis(m3))

# compute percentile interval of mean
sIQ.seq <- seq( from=-3 , to=3 , length.out=30 )
mu <- link( m3 , data=data.frame(SocialIQ = sIQ.seq) )
mu.PI <- apply( mu , 2 , PI )
# plot it all
plot( ADOS ~ SocialIQ , data=d_sub.s , col=rangi2 )
abline( m3 )
shade( mu.PI , sIQ.seq )

```



2. Do the different aspects of IQ account for different portions of the variance in ADOS? 
2.1. Does it make sense to have all IQ measures in the same model? First write a few lines answering the question and motivating your answer, including a discussion as to what happens when you put all of them in the same model. Then build a model following your answer. If your answer is "no", you are not free, you still have to answer: are there alternative ways of answering the question?
2.2. Build the model, assess its quality, write a few lines interpreting the results.
```{r}
# Correlation
# Pairs plot - to eyeball possible correlation between variables
pairs( ~ ADOS + VerbalIQ + NonVerbalIQ + SocialIQ ,
       data=d_sub.s , col=rangi2 )
# Correlation
cor( d_sub.s$NonVerbalIQ , d_sub.s$VerbalIQ )
cor( d_sub.s$NonVerbalIQ, d_sub.s$SocialIQ)
cor( d_sub.s$VerbalIQ, d_sub.s$SocialIQ)

# Multivariate model ADOS ~ VerbalIQ + NonVerbalIQ + SocialIQ
m4 <- map(
  alist(
    ADOS ~ dnorm( mu , sigma ) ,
    mu <- a + bv*VerbalIQ + bn*NonVerbalIQ + bs*SocialIQ ,
    a ~ dnorm( 0 , 0.5 ) ,
    bv ~ dnorm( 0 , 0.5 ) ,
    bn ~ dnorm( 0 , 0.5 ) ,
    bs ~ dnorm( 0 , 0.5 ) ,
    sigma ~ dunif( 0 , 2 )
  ) ,
  data=d_sub.s )
precis(m4, corr=T)

precis_plot(precis(m4))


```


3. Let's now include also the TD children. Does it make sense to ask whether IQ and ADOS are related? Motivate your answer. In any case, if you wanted to build a model to answer that question, which model would you build? Run the model, assess its quality, write a few lines interpreting the results.
```{r}
# How many TDs have ADOS=0? How many ASDs?
sum(d2$ADOS[d2$ASD == 0] == 0)/length(d2$ADOS[d2$ASD == 0]) # TDs
summary(d2$ADOS[d2$ASD == 0])
sd(d2$ADOS[d2$ASD == 0])

sum(d2$ADOS[d2$ASD == 1] == 0)/length(d2$ADOS[d2$ASD == 1]) # ASDs
summary(d2$ADOS[d2$ASD == 1])
sd(d2$ADOS[d2$ASD == 1])



# Multivariate model - using both ASDs and TDs (but still only visit 1)
m5 <- map(
  alist(
    ADOS ~ dnorm( mu , sigma ) ,
    mu <- a + bv*VerbalIQ + bn*NonVerbalIQ + bs*SocialIQ ,
    a ~ dnorm( 0 , 0.5 ) ,
    bv ~ dnorm( 0 , 0.5 ) ,
    bn ~ dnorm( 0 , 0.5 ) ,
    bs ~ dnorm( 0 , 0.5 ) ,
    sigma ~ dunif( 0 , 2 )
  ) ,
  data=d2.s)
precis(m5)
precis_plot(precis(m5))


# Plots

# Only ASDs model
mu <- link( m4 )
# summarize samples across cases
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI )
# simulate observations
# again no new data, so uses original data
ADOS.sim <- sim( m4 , n=1e4 )
ADOS.HPDI <- apply( ADOS.sim , 2 , HPDI )

#plotting
plot( mu.mean ~ d_sub.s$ADOS , col=rangi2 , ylim=range(mu.HPDI) ,
xlab="Observed ADOS" , ylab="Predicted ADOS" )
abline( a = 0 , b = 1, lty = 2)
for ( i in 1:nrow(d_sub.s) )
lines( rep(d_sub.s$ADOS[i],2) , c(mu.HPDI[1,i],mu.HPDI[2,i]) ,
col=rangi2 )

withoutTDp <- recordPlot()


# ASDs + TDs model
mu <- link( m5 )
# summarize samples across cases
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI )
# simulate observations
# again no new data, so uses original data
ADOS.sim <- sim( m5 , n=1e4 )
ADOS.HPDI <- apply( ADOS.sim , 2 , HPDI )

#plotting
plot( mu.mean ~ d2.s$ADOS , col=rangi2 , ylim=range(mu.HPDI) ,
xlab="Observed ADOS" , ylab="Predicted ADOS" )
abline( a = 0 , b = 1, lty = 2)
for ( i in 1:nrow(d2.s) )
lines( rep(d2.s$ADOS[i],2) , c(mu.HPDI[1,i],mu.HPDI[2,i]) ,
col=rangi2 )

withTDp <- recordPlot()

  

# Including diagnosis into the model
m6 <- map( 
  alist(
    ADOS ~ dnorm(mu, sigma),
    mu <- a + bv*VerbalIQ + bn*NonVerbalIQ + bs*SocialIQ + bd*ASD,
    a ~ dnorm( 0 , 0.5 ) ,
    bv ~ dnorm( 0 , 0.5 ) ,
    bn ~ dnorm( 0 , 0.5 ) ,
    bs ~ dnorm( 0 , 0.5 ) ,
    bd ~ dnorm(1, 0.5),
    sigma ~ dunif( 0 , 2 ) 
  ), data = d2.s)
precis(m6)
precis_plot(precis(m6))


# Plot
mu <- link( m6 )
# summarize samples across cases
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI )
# simulate observations
# again no new data, so uses original data
ADOS.sim <- sim( m6 , n=1e4 )
ADOS.HPDI <- apply( ADOS.sim , 2 , HPDI )

#plotting
plot( mu.mean ~ d2.s$ADOS , col=rangi2 , ylim=range(mu.HPDI) ,
xlab="Observed ADOS" , ylab="Predicted ADOS" )
abline( a = 0 , b = 1, lty = 2)
for ( i in 1:nrow(d2.s) )
lines( rep(d2.s$ADOS[i],2) , c(mu.HPDI[1,i],mu.HPDI[2,i]) ,
col=rangi2 )

withTDp <- recordPlot()


# Including interaction between the diagnosis and IQ measures
m7 <- map( 
  alist(
    ADOS ~ dnorm(mu, sigma),
    mu <- a + bn*NonVerbalIQ + bv*VerbalIQ + bs*SocialIQ + bd * ASD + b1 * ASD * NonVerbalIQ + b2 *ASD * VerbalIQ + b3 * ASD * SocialIQ ,
    c(a, bs, bv, bn, b1, b2, b3) ~ dnorm(0, 0.5), 
    bd ~ dnorm(1, 0.5),
    sigma ~ dunif(0,2) 
  ), data = d2.s)
precis(m7)
precis_plot(precis(m7))


# Plot
mu <- link( m7 )
# summarize samples across cases
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI )
# simulate observations
# again no new data, so uses original data
ADOS.sim <- sim( m7 , n=1e4 )
ADOS.HPDI <- apply( ADOS.sim , 2 , HPDI )

#plotting
plot( mu.mean ~ d2.s$ADOS , col=rangi2 , ylim=range(mu.HPDI) ,
xlab="Observed ADOS" , ylab="Predicted ADOS" )
abline( a = 0 , b = 1, lty = 2)
for ( i in 1:nrow(d2.s) )
lines( rep(d2.s$ADOS[i],2) , c(mu.HPDI[1,i],mu.HPDI[2,i]) ,
col=rangi2 )

withTDp <- recordPlot()


# Variatiability in ADOS scores in TDs and ASDs
d2$ASD=as.factor(d2$ASD)

ggplot() + 
  geom_density(data=d2, aes(x=ADOS, group=ASD, color=ASD)) + 
  xlab("ADOS") +
  ylab("Density")+
  theme_classic()


```


4. Let's discuss contents:
4.1. You have three scores for IQ, do they show shared variance? Is that the same in TD and ASD? What does that tell us about IQ?
4.2. You have explored the relation between IQ and ADOS. How do you explain that relation from a cognitive perspective? N.B. You can present alternative hypotheses.
```{r}
# Shared variance (correlation) - ASD only
# Pairs plot - to eyeball possible correlation between variables
pairs( ~ ADOS + VerbalIQ + NonVerbalIQ + SocialIQ ,
       data=d_sub.s , col=rangi2 )
# Correlation
cor( d_sub.s$NonVerbalIQ , d_sub.s$VerbalIQ )
cor( d_sub.s$NonVerbalIQ, d_sub.s$SocialIQ)
cor( d_sub.s$VerbalIQ, d_sub.s$SocialIQ)


# Shared variance (correlation) - TD only
# Pairs plot 
pairs( ~ ADOS + VerbalIQ + NonVerbalIQ + SocialIQ ,
       data=d3.s , col=rangi2 )

# Correlation
cor( d3.s$NonVerbalIQ , d3.s$VerbalIQ )
cor( d3.s$NonVerbalIQ, d3.s$SocialIQ)
cor( d3.s$VerbalIQ, d3.s$SocialIQ)


# Boxplots
d2$ASD = factor(d2$ASD, levels = c("0", "1"), labels = c("TD", "ASD"))
d2 = plyr::rename(d2, c(ASD="Diagnosis"))


ggplot(d2, aes(Diagnosis, NonVerbalIQ)) + 
  geom_boxplot() 
  #facet_wrap(~ASD, labeller = label_both) 

ggplot(d2, aes(Diagnosis, VerbalIQ)) + 
  geom_boxplot() 

ggplot(d2, aes(Diagnosis, SocialIQ)) + 
  geom_boxplot() 

```




5. Bonus questions: Including measurement errors. 
5.1. Let's assume that ADOS has a measurement error of 1. How would you model that in any of the previous models? 
5.2. We know that IQ has an estimated measurement error of 2.12. How would you include that? 


